---
title: "Spotify Analysis Final Project"
author: "Fareeha Ahmad"
date: "2026-06-23"
output: html_document
---

## R Markdown
------------------------------------------------------------------------------------
## Introduction
In this project I will analyze a dataset of ~30000 Spotify songs to see what 
characteristics of songs relate to its popularity. I will use these characteristics 
to predict whether or not a song will chart.

Dataset: https://www.kaggle.com/datasets/joebeachcapital/30000-spotify-songs 

## Loading in data from spotify_songs.csv and calling it spotify
```{r}
spotify = read.csv("spotify_songs.csv")
```

## Going to start by inspecting the data to see what needs to be cleaned
```{r}
# Inspecting the data
str(spotify)
names(spotify)
nrow(spotify)
ncol(spotify)

```

# Cleaning the data
```{r}
# Checking for any missing values in the database
sum(complete.cases(spotify)) 
nrow(spotify)
# 5 less complete rows than total rows
```

```{r}
# Checking if specific columns have missing values
sum(is.na(spotify$danceability))
sum(is.na(spotify$energy))
sum(is.na(spotify$instrumentalness))
sum(is.na(spotify$track_popularity))
# No missing values in these rows
```

```{r}
# Removing rows with missing values
spotify = na.omit(spotify)
# Using these next two queries to see if the number of complete rows matches the number of rows total
nrow(spotify) 
sum(complete.cases(spotify)) 
# Rows match up now
```

```{r}
# Confirming data types that need to be changed
class(spotify$playlist_genre)
class(spotify$key)
class(spotify$mode)

# Make genre, mode, and key factors
spotify$playlist_genre = as.factor(spotify$playlist_genre)
spotify$key = as.factor(spotify$key)
spotify$mode = as.factor(spotify$mode)

# Recheck structure to confirm changes
str(spotify)
```

## Now I am going to create a charted variable based on track_popularity column being over 70
```{r}
spotify$charted = "No" # Creating the column and filling it with No
spotify$charted[spotify$track_popularity >= 70] = "Yes" # Making it Yes if it is over 70
spotify$charted = as.factor(spotify$charted) # Also has to be a factor

table(spotify$charted) # To see how many charted variables there are
```

## Descrptive Analysis
```{r}
# Summary statistics fro the whole dataframe
summary(spotify)
```
```{r}
# Summary statistics for specific numeric variables
summary(spotify$danceability)
sd(spotify$danceability)

summary(spotify$energy)
sd(spotify$energy)

summary(spotify$track_popularity)
sd(spotify$track_popularity)

summary(spotify$instrumentalness)
sd(spotify$instrumentalness)

summary(spotify$acousticness)
sd(spotify$acousticness)
```
```{r}
# Frequency tables for factor variables
table(spotify$playlist_genre)
table(spotify$key)
table(spotify$mode)
```

```{r}
# Frequency tables of songs that charted, and charted songs with the genres they are in
table(spotify$charted)
table(spotify$charted, spotify$playlist_genre)
```

```{r}
# Numeric variable averages by genre
tapply(spotify$danceability, spotify$playlist_genre , mean)
tapply(spotify$energy, spotify$playlist_genre , mean)
tapply(spotify$track_popularity, spotify$playlist_genre , mean)
```

```{r}
# Correlation of every relevant numeric feature against track popularity, this will help decide what to use in the predictive model

cor(spotify$danceability, spotify$track_popularity)
cor(spotify$energy, spotify$track_popularity) 
cor(spotify$loudness, spotify$track_popularity)
cor(spotify$speechiness, spotify$track_popularity)
cor(spotify$acousticness, spotify$track_popularity)
cor(spotify$instrumentalness, spotify$track_popularity) 
cor(spotify$liveness, spotify$track_popularity)
cor(spotify$valence , spotify$track_popularity)
cor(spotify$tempo , spotify$track_popularity)
cor(spotify$duration_ms , spotify$track_popularity) 

# Highest correlation: Instrumentalness, Energy, Duration (None of the correlations are particularly high)
```

```{r}
# Making the plots (Histograms)
hist(spotify$track_popularity, main = "Track Popularity Distribution", col = "lightblue")

hist(spotify$energy, main = "Energy Distribution", col = "lightgreen")

hist(spotify$instrumentalness, main = "Instrumentalness Distribution", col = "lightpink")

hist(spotify$duration_ms, main = "Duration Distribution", col = "lightyellow")

```

```{r}
# Making the plots (Boxplots)
boxplot(spotify$track_popularity~spotify$charted, horizontal = T, col = "yellowgreen", main = "Track Popularity by Status of Charted")

boxplot(spotify$track_popularity~spotify$playlist_genre, horizontal = T, col = "orange", main = "Popularity by Genre")
```
```{r}
# Making the plots (Scatterplots)
plot(spotify$instrumentalness, spotify$track_popularity, col = "purple", main = "Instrumentalness vs Track Popularity")

plot(spotify$energy, spotify$track_popularity, col = "red", main = "Energy vs Track Popularity")

plot(spotify$duration_ms, spotify$track_popularity, col = "darkgreen", main = "Duration vs Track Popularity")

# These reemphasize the lack of correlation as well
```
## Predictive Analysis

# -------------- Install the packages
install.packages("rpart")
install.packages("rpart.plot")
# ---------------------------------------

```{r}
# Load each library
library(rpart)
library(rpart.plot)
```

```{r}
# Creating a subset from the dataset with 70% of the rows assigned to a train dataframe and the remaining 30% assigned to a test dataframe
smp_size = floor(0.70 * nrow(spotify))
set.seed(123)
train_ind = sample(seq_len(nrow(spotify)), size = smp_size)
train = spotify[train_ind, ]
test = spotify[-train_ind, ]
```

```{r}
# Run the decision tree function, this will generate the algorithm to predict charted status based on the characteristics of the song
spotify_model = rpart(charted~instrumentalness+energy+duration_ms, train, method = "class")
```

```{r}
# Displaying the decision tree 
rpart.plot(spotify_model, type = 4, extra = 101)

# Decision tree collapses into just one node, there are no splits because none of the predictors are strong enough variables to predict charting
```

```{r}
table(test$charted)
# Test set is imbalanced
```

```{r}
p = predict(spotify_model, test, type = "class")
```

```{r}
test$predict = p
```


```{r}
table(test[ , c("charted", "predict")])
# Decision tree predicted no for every single song
```
# Checking significancy
- My predictive model showed that the variables I used cannot be used to predict whether or not a song will chart, so I want to check their significance 
```{r}
options(scipen = 999)

R1 = lm(track_popularity~energy, data = spotify)

summary(R1) # This is significant
```

```{r}
options(scipen = 999)

R2 = lm(track_popularity~instrumentalness, data = spotify)

summary(R2) # This is significant
```

```{r}
options(scipen = 999)

R3 = lm(track_popularity~duration_ms, data = spotify)

summary(R3) # This is significant
```

## Prescrptive Analysis
Scenario is that a record label needs to decide how to allocate their limited studio hours to either energy focused tracks or instrumental focused tracks each month, in order to maximize the total popularity of the tracks that they produce

# Step 1: LP formulation
Variables: 
E = energy focused tracks
I = instrumental focused tracks

LP Problem - Objective Function
Max Z = (average popularity of energy tracks)E + (average popularity of instrumental tracks)I
MaxZ = energyavgpop(E) + instavgpop(I)

Subject to (constraints)
4E + 6I = 160    Studio hours
500E + 300I = 12000.     Budget
E >= 0     I >= 0      Non-negativity          

```{r}
# Load lpSolve
library(lpSolve)
```

```{r}
# Average popularity numbers from data that will be used as objective coefficients
energyavgpop = mean(spotify$track_popularity[spotify$energy > 0.7])
instavgpop = mean(spotify$track_popularity[spotify$instrumentalness > 0.5])

energyavgpop
instavgpop
# Checking the values
```

```{r}
# Objective: Vector
objective = c(energyavgpop, instavgpop)
```

```{r}
# Constr: matrix
constr = matrix(c(4,6,500,300), nrow = 2, byrow = TRUE)
```


```{r}
# Constr.dir: Vector
constr.dir = c("<=", "<=")
```


```{r}
# Rhs: Vector
rhs = c(160, 12000)
```


```{r}
# Find the optimal solution
optimum = lp(direction = "max", objective, constr, constr.dir, rhs, compute.sens = TRUE)
```


```{r}
# Display the optimum solution values
optimum$solution
# E: 13.33333
# I: 17.77778
```

```{r}
# Display optimum objective value 
optimum$objval
```


```{r}
# Obtain dual values 
optimum$duals
# Studio hours: 2.1009426
# Budget:0.0640122
# E: 0.0000000
# I: 0.0000000
```

## Knit to HTML